It's all about Meaning!\\{\normalsize A Theory for Data Migration during Software Development

### 2022-12-11 2022 Migration/articleMigration.tex
fixed the signature function

   The relation $\inst:\Pair{\Atoms}{\Concepts}$ is the instance relation between atoms and concepts.
   The expression $a\ \inst\ C$ means that atom $a$ is an \emph{instance} of concept $C$.
   This relation is used in the type system, in which $\inst$ assigns a type to every atom.
   This makes $\inst$ total on the atoms in the dataset:
\begin{eqnarray}
   \forall a\in\atoms\ \exists C\in\Concepts: a\ \inst\ C
\label{eqn:typedAtom}
\end{eqnarray}

@@

   Triples serve to represent data.
   A triple is an element of $\Triple{\Atoms}{\Rels}{\Atoms}$.
   A dataset $\dataset$ is a tuple $\pair{\triples}{\inst}$ that satisfies:
\begin{eqnarray}
   \forall\triple{a}{\declare{n}{A}{B}}{b}\in\triples&:&a\ \inst\ A\ \wedge\ b\ \inst\ B
   \label{eqn:wellTypedEdge}
\end{eqnarray}
   For example, $\triple{\text{\atom{Peter}}}{\declare{\id{phone}}{\tt Person}{\tt TelephoneNumber}}{\text{\atom{074238991}}}$ is a triple.
   Equation~\ref{eqn:wellTypedEdge} says that \atom{Peter} is an instance of {\tt Person} and \atom{074238991} is an instance of {\tt TelephoneNumber}.
   In practice, users can say that Peter has telephone number 074238991.
   So, the ``thing'' that \atom{Peter} refers to (which is Peter) has \atom{074238991} as a telephone number.
   This ``meaning from practice'' has no consequences in the formal world.
   We leave it entirely up to a user to attach a practical meaning to a triple.

   $\triples$ is a finite set of triples and $\inst$ is the instance relation.
   In case we need to be explicit about the dataset,
   we will write $\triples_{\dataset}$ and $\inst_{\dataset}$ to refer to $\triples$ and $\inst$ respectively.
   Every dataset is an element of an infinite set called $\Dataset$.
   To save writing in the sequel, we will write $a\ r\ b$ to denote that $\triple{a}{r}{b}\in\triples$.

   A relation $r$ can serve as a container of pairs,
   as defined by the function $\id{pop}_r:\Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$.
   It defines a set of pairs, which we call the population of $r$:
\begin{equation}
   \pop{r}{\dataset}\ =\ \{ \pair{a}{b}\mid\ \triple{a}{r}{b}\in\triples_{\dataset}\}
\label{eqn:pop}
\end{equation}
   Equation~\ref{eqn:wellTypedEdge} implies that for every dataset $\dataset$:
\[\pair{a}{b}\in\pop{\declare{n}{A}{B}}{\dataset}\ \Rightarrow\ \inst\ A\ \wedge\ b\ \inst\ B\]
   For a developer, this means that the type of an atom depends only on the relation in which it resides; not on the actual population of the database.
   This allows for static typing, which has well established advantages for the software engineering process~\cite{HanenbergKRTS14,Petersen2014}.

@@

   A schema $\schema$ is a triple $\triple{\concepts}{\rels}{\rules}$,
   in which $\concepts$ is a finite set of concepts,
   $\rels$ is a finite set of relations,
   and $\rules$ is a finite set of rules.
   We write $\schema_{\infsys}$ to refer to the schema of information system $\infsys$.
   We write $\concepts_{\schema}$, $\rels_{\schema}$, and $\rules_{\schema}$ to refer to $\concepts$, $\rels$, and $\rules$ respectively.
%  Every schema is an element of an infinite set called $\Schema$.

@@

   A rule in a schema serves to constrain the dataset, to ensure its semantic integrity.
   For every rule $u$ in the schema, we assume there is a function $\id{viol}_u : \Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$
   that represents the set of violations of rule $u$ in any dataset.
   We also assume a function $\Sign : \rules\rightarrow\Pair{\Concepts}{\Concepts}$
   that represents the signature of a rule, i.e. the type of atoms in violations:
\begin{eqnarray}
   \sign{u}=\pair{A}{B}\\
   \pair{a}{b}\in\viol{u}{\dataset}&\Rightarrow&a\ \inst\ A\wedge b\ \inst\ B
   \label{eqn:wellTypedEdged violations}
\end{eqnarray}
   To determine whether rule $u$ is satisfied in dataset $\dataset$,
   we define $\sat{u}{\Dataset}$:
\begin{eqnarray}
   \sat{u}{\dataset}&\Leftrightarrow&\viol{u}{\dataset}=\emptyset
   \label{eqn:sat}
\end{eqnarray}
   Violations are used to produce meaningful error messages and to trigger changes in the dataset to restore satisfaction of that rule.

@@

\subsection{Information Systems}
\label{sct:Information Systems}
   This section defines the notion of information system.
   As before, we will suffix the elements of this definition if the situation requires it.
\begin{definition}[information system]
\label{def:information system}
\item An information system $\infsys$ is a tuple $\la\dataset,\schema,\roles,\maintain\ra$, in which
\begin{itemize}
   \item dataset $\dataset$ is defined as in section~\ref{sct:Datasets};
   \item schema $\schema$ is defined as in section~\ref{sct:Schemas};
   \item $\roles$ is a set of roles;
   \item $\maintain : \roles\times\rules$ is the maintainance relation between roles and rules from $\schema$.
\end{itemize}
\end{definition}
   A \define{role} is a name that identifies a group of users.
   It serves as a placeholder for a person or a machine (i.e. an actor) who can change the dataset.
   The purpose of a role is to mention an individual user (human) or an automated actor (bot) without knowing who that user is.
   In the sequel, when we talk about a role we actually mean an actor who fulfills that role in the system.

@@

   This example contains a dataset with seven triples
   and a schema with three concepts, three relations, and one rule.
   Now let us check the requirements to verify that this example defines an information system.
   The Ampersand compiler generates a dataset $\dataset$, which contains a set of triples and a relation $\inst$.
   It defines the set of triples $\triples$ as:
   \[\begin{array}[t]{l}
         \triple{\text{\tt "Peter"}}{\declare{\text{\tt takes}}{\text{\tt Student}}{\text{\tt Course}}}{\text{\tt "Management"}}\\
         \triple{\text{\tt "Susan"}}{\declare{\text{\tt takes}}{\text{\tt Student}}{\text{\tt Course}}}{\text{\tt "Business IT"}}\\
         \triple{\text{\tt "John"}}{\declare{\text{\tt takes}}{\text{\tt Student}}{\text{\tt Course}}}{\text{\tt "Business IT"}}\\
         \triple{\text{\tt "Finance"}}{\declare{\text{\tt isPartOf}}{\text{\tt Module}}{\text{\tt Course}}}{\text{\tt "Management"}}\\
         \triple{\text{\tt "Business Rules"}}{\declare{\text{\tt isPartOf}}{\text{\tt Module}}{\text{\tt Course}}}{\text{\tt "Business IT"}}\\
         \triple{\text{\tt "Business Analytics"}}{\declare{\text{\tt isPartOf}}{\text{\tt Module}}{\text{\tt Course}}}{\text{\tt "Business IT"}}\\
         \triple{\text{\tt "IT-Governance"}}{\declare{\text{\tt isPartOf}}{\text{\tt Module}}{\text{\tt Course}}}{\text{\tt "Management"}}
   \end{array}\]
   The relation $\id{inst}$ contains the pairs:
   \[\begin{array}{l}
      \pair{\tt "Finance"}{\tt Module}\\
      \pair{\tt "Business Rules"}{\tt Module}\\
      \pair{\tt "Business Analytics"}{\tt Module}\\
      \pair{\tt "IT-Governance"}{\tt Module}\\
      \pair{\tt "Management"}{\tt Course}\\
      \pair{\tt "Business IT"}{\tt Course}\\
      \pair{\tt "Peter"}{\tt Student}\\
      \pair{\tt "Susan"}{\tt Student}\\
      \pair{\tt "John"}{\tt Student}
   \end{array}\]
   Requirement~\ref{eqn:typedAtom} is satisfied because every atom has a concept.
   The tuple $\pair{\triples}{\inst}$ satisfies both requirements~\ref{eqn:wellTypedEdge} and~\ref{eqn:pop}.
   So, this defines a dataset $\dataset=\pair{\triples}{\inst}$ as introduced in section~\ref{sct:Datasets}.

@@

\subsection{Changes}
% The purpose of this section is to explain why a dataset is structured the way it is.
   In the migration of a dataset we deal with changes to the elements that are not data:
   $\inst$, $\concepts$, and $\rels$.
   Such changes have further reaching consequences, however.
   Changes to $\inst$, $\concepts$, $\rels$ change the data structure in a dataset.
   We define a \define{migration of a dataset} $\la\atoms,\concepts,\inst,\rels,\dataset\ra$ as a change in which one or more of $\inst$, $\concepts$, or $\rels$ change.
   To satisfy the definition of datasets, $\dataset$ and $\atoms$ must change too,
   but the migration should try to preserve the maximal amount of data.

>>>>

   The relation $\inst:\Pair{\Atoms}{\Concepts}$ is the instance relation between atoms and concepts.
   The expression $a\ \inst\ C$ means that atom $a$ is an \emph{instance} of concept $C$.
   This relation is used in the type system, in which $\inst$ assigns a type to every atom in the dataset.

@@

   Triples serve to represent data.
   A triple is an element of $\Triple{\Atoms}{\Rels}{\Atoms}$.

   A dataset $\dataset$ is a tuple $\pair{\triples}{\inst}$ that satisfies:
\begin{eqnarray}
   \forall\triple{a}{\declare{n}{A}{B}}{b}\in\triples&:&a\ \inst\ A\ \wedge\ b\ \inst\ B
   \label{eqn:wellTypedEdge}
\end{eqnarray}
   For example, $\triple{\text{\atom{Peter}}}{\declare{\id{phone}}{\tt Person}{\tt TelephoneNumber}}{\text{\atom{074238991}}}$ is a triple.
   Equation~\ref{eqn:wellTypedEdge} says that \atom{Peter} is an instance of {\tt Person} and \atom{074238991} is an instance of {\tt TelephoneNumber}.
   In practice, users can say that Peter has telephone number 074238991.
   So, the ``thing'' that \atom{Peter} refers to (which is Peter) has \atom{074238991} as a telephone number.
   This ``meaning from practice'' has no consequences in the formal world.
   We leave it entirely up to a user to attach a practical meaning to a triple.

   In case we need to be explicit about the dataset,
   we will write $\triples_{\dataset}$ and $\inst_{\dataset}$ to refer to $\triples$ and $\inst$ respectively.
   Every dataset is an element of an infinite set called $\Dataset$.
   To save writing in the sequel, we may write $a\ r\ b$ to denote that $\triple{a}{r}{b}\in\triples$.

   A relation $r$ can serve as a container of pairs,
   as defined by the function $\id{pop}_r:\Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$.
   It defines a set of pairs, which we call the population of $r$:
\begin{equation}
   \pop{r}{\dataset}\ =\ \{ \pair{a}{b}\mid\ \triple{a}{r}{b}\in\triples_{\dataset}\}
\label{eqn:pop}
\end{equation}
   Equation~\ref{eqn:wellTypedEdge} implies that for every dataset $\dataset$:
\[\pair{a}{b}\in\pop{\declare{n}{A}{B}}{\dataset}\ \Rightarrow\ a\ \inst\ A\ \wedge\ b\ \inst\ B\]
   For a developer, this means that the type of an atom depends only on the relation in which it resides; not on the actual population of the database.
   This allows for static typing, which has well established advantages for the software engineering process~\cite{HanenbergKRTS14,Petersen2014}.

@@

   A schema $\schema$ is a triple $\triple{\concepts}{\rels}{\rules}$,
   in which $\concepts$ is a finite set of concepts,
   $\rels$ is a finite set of relations,
   and $\rules$ is a finite set of rules.
   We write $\schema_{\infsys}$ to refer to the schema of information system $\infsys$.
   If needed, we write $\concepts_{\schema}$, $\rels_{\schema}$, and $\rules_{\schema}$ to refer to $\concepts$, $\rels$, and $\rules$ respectively.
%  Every schema is an element of an infinite set called $\Schema$.

@@

   A rule in a schema serves to constrain the dataset, to ensure its semantic integrity.
   Every rule is an element of an infinite set called $\Rules$,
   which is disjoint from $\Concepts$, $\Rels$, and $\Atoms$.
   For every rule $u$ in the schema, we assume there is a function $\id{viol}_u : \Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$
   that represents the set of violations of rule $u$ in any dataset.
   We also assume a function $\id{sign} : \Rules\rightarrow\Pair{\Concepts}{\Concepts}$
   that represents the signature of a rule, i.e. the type of atoms in violations:
\begin{eqnarray}
   \sign{u}=\pair{A}{B}\ \wedge\ \pair{a}{b}\in\viol{u}{\dataset}&\Rightarrow&a\ \inst\ A\wedge b\ \inst\ B
   \label{eqn:wellTypedEdged violations}
\end{eqnarray}
   To determine whether rule $u$ is satisfied in dataset $\dataset$,
   we define $\sat{u}{\Dataset}$:
\begin{eqnarray}
   \sat{u}{\dataset}&\Leftrightarrow&\viol{u}{\dataset}=\emptyset
   \label{eqn:sat}
\end{eqnarray}
   Violations are used to produce meaningful error messages and to trigger changes in the dataset to restore satisfaction of that rule.

@@

\subsection{Information Systems}
\label{sct:Information Systems}
   This section defines the notion of information system.
   As before, we will suffix the elements of this definition if the situation requires it.
\begin{definition}[information system]
\label{def:information system}
\item An information system $\infsys$ is a tuple $\la\dataset,\schema,\roles,\maintain\ra$, in which
\begin{itemize}
   \item dataset $\dataset=\pair{\triples}{\inst}$ is defined as in section~\ref{sct:Datasets};
   \item schema $\schema=\triple{\concepts}{\rels}{\rules}$ is defined as in section~\ref{sct:Schemas};
   \item $\roles$ is a set of roles;
   \item $\maintain : \roles\times\rules$ is the maintainance relation between roles and rules from $\schema$.
\end{itemize}
\end{definition}
   A \define{role} is a name that identifies a group of users.
   It serves as a placeholder for a person or a machine (i.e. an actor) who can change the dataset.
   The purpose of a role is to mention an individual user (human) or an automated actor (bot) without knowing who that user is.
   In the sequel, when we talk about a role we actually mean an actor who fulfills that role in the system.

@@

   This example contains a dataset with seven triples
   and a schema with three concepts, three relations, and one rule.
   Now let us check the requirements to verify that this example defines an information system.
   The Ampersand compiler generates a dataset $\dataset$, which contains a set of triples and a relation $\inst$.
   It defines the set of triples $\triples$ as:
   \[\begin{array}[t]{l}
         \triple{\text{\tt "Peter"}}{\declare{\text{\tt takes}}{\text{\tt Student}}{\text{\tt Course}}}{\text{\tt "Management"}}\\
         \triple{\text{\tt "Susan"}}{\declare{\text{\tt takes}}{\text{\tt Student}}{\text{\tt Course}}}{\text{\tt "Business IT"}}\\
         \triple{\text{\tt "John"}}{\declare{\text{\tt takes}}{\text{\tt Student}}{\text{\tt Course}}}{\text{\tt "Business IT"}}\\
         \triple{\text{\tt "Finance"}}{\declare{\text{\tt isPartOf}}{\text{\tt Module}}{\text{\tt Course}}}{\text{\tt "Management"}}\\
         \triple{\text{\tt "Business Rules"}}{\declare{\text{\tt isPartOf}}{\text{\tt Module}}{\text{\tt Course}}}{\text{\tt "Business IT"}}\\
         \triple{\text{\tt "Business Analytics"}}{\declare{\text{\tt isPartOf}}{\text{\tt Module}}{\text{\tt Course}}}{\text{\tt "Business IT"}}\\
         \triple{\text{\tt "IT-Governance"}}{\declare{\text{\tt isPartOf}}{\text{\tt Module}}{\text{\tt Course}}}{\text{\tt "Management"}}
   \end{array}\]
   The relation $\id{inst}$ contains the pairs:
   \[\begin{array}{l}
      \pair{\tt "Finance"}{\tt Module}\\
      \pair{\tt "Business Rules"}{\tt Module}\\
      \pair{\tt "Business Analytics"}{\tt Module}\\
      \pair{\tt "IT-Governance"}{\tt Module}\\
      \pair{\tt "Management"}{\tt Course}\\
      \pair{\tt "Business IT"}{\tt Course}\\
      \pair{\tt "Peter"}{\tt Student}\\
      \pair{\tt "Susan"}{\tt Student}\\
      \pair{\tt "John"}{\tt Student}
   \end{array}\]
   The tuple $\pair{\triples}{\inst}$ satisfies both requirements~\ref{eqn:wellTypedEdge} and~\ref{eqn:pop}.
   So, this defines a dataset $\dataset=\pair{\triples}{\inst}$ as introduced in section~\ref{sct:Datasets}.

@@

\subsection{Changes}
% The purpose of this section is to explain why a dataset is structured the way it is.
   In the migration of a dataset we deal with changes to the elements that are not data:
   $\inst$, $\concepts$, and $\rels$.
   Such changes have further reaching consequences, however.
   Changes to $\inst$, $\concepts$, $\rels$ change the data structure in a dataset.
   We define a \define{migration of a dataset} $\la\atoms,\concepts,\inst,\rels,\dataset\ra$ as a change in which one or more of $\inst$, $\concepts$, or $\rels$ change.
   To satisfy the definition of datasets, $\dataset$ must change too,
   but the migration should try to preserve the maximal amount of data.
